---
title: "Downloading SVI Data"
output: html_document
date: "2025-07-03"
editor_options: 
  markdown: 
    wrap: 72
---

```{r set.knitr.options, include= FALSE, message=FALSE, warning=FALSE}
#-----set knitr options-----

# clear environment
rm(list = ls())

# knitr options
suppressMessages(library(knitr))
opts_chunk$set(tidy = FALSE, echo = TRUE, message = FALSE)

# Load pacman, installing if needed and other packages as needed
if (!require("pacman")) { install.packages("pacman", repos = "http://cran.r-project.org") }

# Load other packages, installing as needed.
pacman::p_load(tidyverse, tigris, sf, tmap)

# Set working directory 
setwd("~/Library/Mobile Documents/com~apple~CloudDocs/Documents/SFSU Faculty/PINC 2025/PSP2025/Climate Change")
```

# About SVI Data

The Social Vulnerability Index (SVI), developed by the CDC/ATSDR,
measures the resilience of communities when confronted by external
stresses such as natural disasters or disease outbreaks. It is based on
16 U.S. Census variables grouped into four themes: socioeconomic status,
household composition & disability, minority status & language, and
housing type & transportation. The SVI is reported at the census tract
level and ranges from 0 (least vulnerable) to 1 (most vulnerable),
allowing for comparison across geographic areas.

In our analysis, we use the 2020 SVI to assess how extreme heat exposure
correlates with social vulnerability across San Francisco census tracts
from 2000 to 2020. By linking weekly heat metrics derived from Daymet
data to the SVI scores of each tract, we examine whether communities
with higher social vulnerability experience more frequent or intense
extreme heat events. This helps us identify environmental justice
patterns and prioritize areas for future public health interventions.

The variable *RPL_THEMES* represents the overall percentile rank of
social vulnerability for each census tract, based on the combination of
all four SVI themes: Socioeconomic status Household composition &
disability Minority status & language Housing type & transportation

## Downloading Data

Since no specific API exists, we can directly download from the website:
<https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html>

You can select Year: 2020, Geography: California, Geography Type: Census
Tracts.

```{r}

# Step 0: Load SVI and pre-processed clean daymet data
svi_data <- read_csv("Data/CA_SVI.csv")
# View(svi_data)

sf_daymet <- read_csv("Data/sf_daymet_clean.csv")
# View(sf_daymet)

```

```{r analysis, echo= FALSE}

# Calculate 97.5th Percentile tmax per Census Tract
tmax_thresholds <- sf_daymet %>%
  filter(measurement == "tmax") %>%
  group_by(site) %>%
  summarize(tmax_97.5 = quantile(value, 0.975, na.rm = TRUE))


# Tag Days as EHEs if tmax > 97.5th Percentile
tmax_ehe <- sf_daymet %>%
  filter(measurement == "tmax") %>%
  left_join(tmax_thresholds, by = "site") %>%
  mutate(ehe_day = value > tmax_97.5)


# Convert Day-of-Year to Date, then to Week
tmax_ehe <- tmax_ehe %>%
  mutate(
    date = as.Date(yday - 1, origin = paste0(year, "-01-01")),
    week = isoweek(date),
    year = year(date)
  )

# Summarize Weekly EHE Counts
weekly_ehe <- tmax_ehe %>%
  group_by(site, year, week) %>%
  summarize(
    weekly_ehe_days = sum(ehe_day, na.rm = TRUE),
    avg_weekly_tmax = mean(value, na.rm = TRUE),
    .groups = "drop"
  )

# Join with SVI Data

# since the svi_data has a FIPS code (as a character) with a leading 0, but the tmax_ehe has site (as a double) without the leading 0, data wrangling is needed...
# we can fix this by converting site to character with leading zeroes

weekly_ehe <- weekly_ehe %>%
  mutate(site = as.character(sprintf("%011.0f", site)))


weekly_ehe_svi <- weekly_ehe %>%
  left_join(svi_data, by = c("site" = "FIPS"))

```

```{r}
ehe_summary <- weekly_ehe_svi %>%
  group_by(site) %>%
  summarize(
    total_ehe_days = sum(weekly_ehe_days, na.rm = TRUE),
    mean_ehe_days  = mean(weekly_ehe_days, na.rm = TRUE),
    svi            = first(RPL_THEMES)  # assume SVI doesn't vary weekly
  )

```

```{r}

ggplot(ehe_summary, aes(x = svi, y = mean_ehe_days)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(
    title = "Relationship between Social Vulnerability and Extreme Heat Events",
    x = "SVI (RPL_THEMES)",
    y = "Mean Weekly EHE Days"
  )

```

```{r}
# Spearman correlation
cor.test(ehe_summary$svi, ehe_summary$mean_ehe_days, method = "spearman")

# Pearson correlation
cor.test(ehe_summary$svi, ehe_summary$mean_ehe_days, method = "pearson")

```

```{r}

# Create SVI Quartiles (fixed per tract)
joined_data <- weekly_ehe_svi %>%
  mutate(
    svi_quartile = ntile(RPL_THEMES, 4)  # Q1 = least vulnerable, Q4 = most
  )

# Aggregate to Tract-Year Level
ehe_yearly <- joined_data %>%
  group_by(site, year, svi_quartile) %>%
  summarize(
    yearly_ehe_days = sum(weekly_ehe_days, na.rm = TRUE),
    .groups = "drop"
  )

# Summarize Across Tracts Within Each Quartile-Year
ehe_trends <- ehe_yearly %>%
  group_by(year, svi_quartile) %>%
  summarize(
    mean_ehe = mean(yearly_ehe_days, na.rm = TRUE),
    sd_ehe = sd(yearly_ehe_days, na.rm = TRUE),
    n = n(),
    se_ehe = sd_ehe / sqrt(n),
    .groups = "drop"
  )

```

```{r}
ggplot(ehe_trends, aes(x = year, y = mean_ehe, color = factor(svi_quartile))) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = mean_ehe - se_ehe, ymax = mean_ehe + se_ehe, fill = factor(svi_quartile)), alpha = 0.2, color = NA) +
  scale_color_brewer(palette = "YlOrRd", name = "SVI Quartile") +
  scale_fill_brewer(palette = "YlOrRd", name = "SVI Quartile") +
  labs(
    title = "Trends in Annual Extreme Heat Events by Social Vulnerability",
    x = "Year",
    y = "Mean EHE Days per Tract"
  ) +
  theme_minimal()
```

```{r, message = FALSE}

# Get the San Francisco County Boundary
sf_boundary <- counties(state = "CA", year = 2020, class = "sf") %>%
  filter(NAME == "San Francisco") %>%
  st_transform(4326)  # ensure consistent CRS

# Download Tracts for San Francisco
tracts_sf <- tracts(state = "CA", county = "San Francisco", year = 2020, class = "sf") %>%
  st_transform(4326)

# Keep Only Tracts That Touch SF Land
tracts_sf_clean <- tracts_sf[sf::st_intersects(tracts_sf, sf_boundary, sparse = FALSE), ]

# Join with EHE + SVI Data
ehe_map_data <- ehe_map_data %>%
  mutate(site = as.character(site))

tracts_sf_data <- tracts_sf_clean %>%
  left_join(ehe_map_data, by = c("GEOID" = "site"))

```

## Data Preparation

Join to weekly heat exposure data using tract ID (FIPS)

Exploratory Data Analysis

4.1 Time trends Plot yearly average heat days per tract

Plot number of heat waves per year

4.2 Spatial patterns Map of average weekly heat days per tract

Map of SVI values per tract

Map showing both for visual overlap

Statistical Analysis

5.1 Correlation analysis Correlation between average weekly heat
exposure and SVI

Include Pearson or Spearman correlation and scatterplot

5.2 Group comparison Create SVI categories (quartiles or high/low)

Compare average heat exposure across SVI groups (boxplots, t-tests or
Kruskal-Wallis)

```{r session.info, echo=FALSE}
#-----session info-----

sessionInfo()
```
